---
title: Seasonal and annual trends in cyanotoxins within eastern Nebraskan lakes
subtitle: A case study from three hydromorphically different lakes
output: 
  pdf_document:
    highlight: tango
    fig_caption: yes
    number_sections: yes
geometry: margin=0.75in
fontsize: 11pt

---
**Megan L. Larsen** and Daniel D. Snow 

*University of Nebraska-Lincoln Water Sciences Laboratory*

1840 N. 37th Street

Lincoln, Nebraska USA 68583-0844

This analysis has been prepared in conjunction with our manuscript for THIS JOURNAL

Data analysis for this project was completed with `r sessionInfo()$R.version$version.string`.

\clearpage


#Project Summary

##**Overview**


##**Project questions**

What are the temporal (annual and seasonal) patterns of microcystins in the selected lake/reservoir systems? 
-- Focus on microcystins with coupled lake monitoring data available in Storet

What are the temporal patterns in physical, chemical, and biological characteristics of Nebraska Lakes? 
-- Incorporate information from the Ambient Lake Monitoring programs with NDEQ with microcystins 
-- make note of lake treatments

How do different analytical methods (ELISA vs. LC/MS) compare for the detection of algal toxins? 
-- Total concentration 
-- Specificity (ELISA = broad; LC/MS = analyte specific) 
-- detection of other cyanotoxins: Many programs have regulations for other toxins (Ohio)

##**List of abbrevations**

NDEQ: Nebraska Department of Environmental Quality

USEPA: United States Environmental Protection Agency

USACOE: United States Army Corp of Engineers

USGS: United States Geological Survey

NOAA: National Oceanic and Atmospheric Administration

\newpage 
\tableofcontents 
\newpage
\listoftables
\newpage
\listoffigures
\newpage


```{r WorkingEnv, echo = FALSE, message=FALSE}
setwd("C:/Users/meglarse/GitHub/ToxTrends/")

knitr::opts_chunk$set(fig.width=7, fig.height=8,
                      echo = FALSE, warning = FALSE, message = FALSE)

require(ggplot2);require(gridExtra);require(pander);require(reshape2)


theme_std <- function (base_size = 12, base_family = "") 
{
    theme_grey(base_size = base_size, base_family = base_family) %+replace% 
        theme(axis.text = element_text(size = rel(0.8)), 
              axis.ticks = element_line(colour = "black"), 
              legend.key = element_rect(colour = "grey80"), 
              panel.background = element_rect(fill = "white", colour = NA), 
              panel.border = element_rect(fill = NA, colour = "grey50"), 
              panel.grid.major = element_line(NA), 
              panel.grid.minor = element_line(NA), 
              strip.background = element_rect(fill = "grey80", colour = "grey50", size = 0.2))
}

theme_set(theme_std())

```

# Materials and Methods
## Lake descriptions
```{r}
require(ggmap)
# Load sample points
ndeq.stn <- read.csv("./data/ndeq_station.csv", header = TRUE)
smp.points <- read.csv("./data/sitedescriptions.csv")
noaa.dat <- read.csv("./data/noaa_data.csv", header = TRUE)

#load libraries for chunk
library(maps);library(mapdata)
library(maptools) # for shapefiles
#library(scales)   #for transparency
#library(mapplots)
#library(rgdal)

library(raster)

# Main Map
main.map <- get_map(location = c(-96.318667,41.084526), zoom = 9,
                       maptype = "toner", source = "stamen")

main <- ggmap(main.map, extent = "panel",legend = "right") +
  geom_point(data = smp.points, aes(LongitudeMeasure, LatitudeMeasure), 
             col = "grey",bg = "red", pch = 21, cex =4) +
  geom_point(data = noaa.dat, aes(x = LONGITUDE, y = LATITUDE), 
             pch = 22, col = "grey", bg = "black", cex = 4) +
  xlab("Longitude") + ylab("Latitude") +
  theme(panel.border = element_rect(fill = NA))
#print(main)

#ggsave("./supporting-files/FigS1-Main.pdf", main, units = "in", 
#       scale = 1.5, height = 4, width = 4)
```

Data for this project was collected from publically available datasets available through the National Oceanic and Atmospheric Administration and the Environmental Protection Agency's STORET download tool. 
Contributing agencies, the data type, and sampling station location are described in Table 1 below.

```{r ListofStations}
stations.table <- smp.points[,c("OrganizationFormalName","LakeName","MonitoringLocationIdentifier","LatitudeMeasure","LongitudeMeasure")]
colnames(stations.table)<-c("Organization","Lake","Location ID","Latitude","Longitude")
pander(stations.table)
```

Prior to the start of any comparisons within or across lakes, we first cleaned-up the dataset by ensuring parameter units were equivalent... The ELISA -adda kit has a lower reporting limit of 0.15 therefore any reported value less than 0.15 \mu g/L were converted to 0 \mu g/L 

```{r dataconvert}
dat <- read.csv("./data/result.csv", header = TRUE)

# Do some data cleanup for this analysis

dat$ResultMeasureValue <- as.numeric(as.character(dat$ResultMeasureValue))

## Convert ActivityStartDate to a Juilan value to look at annual and decadal patterns
require(lubridate)
require(ggplot2)

dat$tmp <- as.Date(dat$ActivityStartDate, format = "%Y-%m-%d") #Converts data to date structure
dat$yr <- as.numeric(format(dat$tmp,'%Y'))         #Time is 
dat$yr.fac <- as.factor(as.numeric(format(dat$tmp,'%Y')))         #Time is now numeric
dat$mo <- as.factor(format(dat$tmp,'%m'))
dat$jul.dat <- yday(dat$tmp)                                  #Time is now in Julian
dat$wk <- week(dat$tmp)

dat <- dat[dat$ActivityMediaName != "Sediment",]
```

```{r decadal-trends}
## Subset the microcystins data

bmp <- dat[dat$CharacteristicName == "Microcystin" & 
             dat$OrganizationIdentifier == "21NEB001_WQX",]

## NDEQ sites of interest
sites <- c("21NEB001_WQX-LLP1FRMT2004","21NEB001_WQX-LLP1FRMT2005","21NEB001_WQX-LLP2PAWNEE03","21NEB001_WQX-LLP2PAWNEE05", "21NEB001_WQX-LMT1CARTER03")

bmp <- bmp[bmp$MonitoringLocationIdentifier %in% sites,]

# Remove quality control samples
bmp <- bmp[bmp$ActivityTypeCode != "Quality Control Sample-Lab Blank" & bmp$ActivityTypeCode != "Quality Control Sample-Field Replicate",]

# Subset for the 2005-2015 decade
bmp <- bmp[bmp$yr != 2004,]

# detection limit is 0.15, change all values less than 0.15 to 0
bmp$ResultMeasureValue[bmp$ResultMeasureValue <0.15] <- 0
```


```{r, include = FALSE}
# Inset Map
pdf("./supporting-files/FigS1-Inset.pdf")
map("state", col = "grey", boundary = TRUE,  lty = 2)
map("worldHires","USA", xlim = c(-130,-65), ylim = c(25,50), 
    col = "black", boundary = TRUE, add = TRUE)
map("state","Nebraska", col = "grey", fill = TRUE, add = TRUE) 
points(x=-96.318667,y=41.084526, pch = 15, col = "red")
dev.off()
```

\clearpage

# Results: Microcystins
## Annual and seasonal trends in cyanotoxins from NDEQ data


```{r, fig.cap = "Annual variation in microcystin concentrations across focal lakes from 2005 - 2015. Microcystin concentrations were quaitified by the NDEQ as part of the Beach Monitoring Program (BMP) using ELISA.", fig.height=6}

p1 = ggplot(bmp, aes(yr.fac,ResultMeasureValue)) +
  geom_hline(yintercept = 20, col = "red", lty = 1, lwd = 1) +
  geom_point(cex = 1) +
  geom_boxplot(fill = NA) + 
  xlab("Year") + ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  facet_wrap(~MonitoringLocationIdentifier, ncol = 2)

print(p1)


#ggsave("./supporting-files/20160818-AnnualTrends.pdf", p1, 
#       scale = 1.5, width = 6, height = 6, units = "in", dpi = 300)
```

\clearpage

```{r IndvLakeTrends, fig.cap= "Annual mean microcystin concentration for each lake system. Lake closure levels for the state of Nebraska prior to 2017 required a toxin concentration > 20 ug/L (red). Effective for the 2017 season, lake closure levels were reduced to 4 ug/L (black)"}
bmp.paw <- bmp[grep("PAW", bmp$MonitoringLocationIdentifier),]
bmp.fre <- bmp[grep("FRM", bmp$MonitoringLocationIdentifier),]
bmp.car <- bmp[grep("CAR", bmp$MonitoringLocationIdentifier),]


annual.paw = ggplot(bmp.paw, aes(yr,ResultMeasureValue)) +
  ylim(0,80) +
  scale_x_continuous(breaks = c(2005:2015)) +
  geom_hline(yintercept = 20, col = "red", lty = 1, lwd = 1) +
  geom_hline(yintercept = 4, col = "black", lty = 1, lwd = 1) +
  geom_jitter(cex = 1, col = "grey50", width = 0.2, height = 0.1)  +
  stat_summary(fun.data = "mean_cl_boot", color = "black", size = 1) +
  xlab("Year") + ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  annotate("text", y = Inf, x = Inf, label = "Pawnee Lake", vjust = 2, hjust = 1.25)

#print(annual.paw)

annual.fre = ggplot(bmp.fre, aes(yr,ResultMeasureValue)) +
  ylim(0,80) +
  scale_x_continuous(breaks = c(2005:2015)) +
  geom_hline(yintercept = 20, col = "red", lty = 1, lwd = 1) +
  geom_hline(yintercept = 4, col = "black", lty = 1, lwd = 1) +
  geom_jitter(cex = 1, col = "grey50", width = 0.2, height = 0.1)  +
  stat_summary(fun.data = "mean_cl_boot", color = "black", size = 1) +
  xlab("Year") + ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  annotate("text", y = Inf, x = Inf, label = "Fremont Lake No. 20", vjust = 2, hjust = 1.15)

#print(annual.fre)

annual.car = ggplot(bmp.car, aes(yr,ResultMeasureValue)) +
  ylim(0,50) +
  scale_x_continuous(breaks = c(2005:2015)) +
  geom_hline(yintercept = 20, col = "red", lty = 1, lwd = 1) +
  geom_hline(yintercept = 4, col = "black", lty = 1, lwd = 1) +
  geom_jitter(cex = 1, col = "grey50", width = 0.2, height = 0.1) +
  stat_summary(fun.data = "mean_cl_boot", color = "black", size = 1) +
  xlab("Year") + ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  annotate("text", y = Inf, x = Inf, label = "Carter Lake", vjust = 2, hjust = 1.25)

#print(annual.car)

grid.arrange(annual.paw,annual.fre, annual.car, ncol = 1)


#Notes:
# -- Add lake labels to plots
# -- add polygons to highlight study years

```

```{r TrendStats}

#H0: Mic concentrations were the same between years?
#H1: Some years had higher concentrations of microcystins.
#H2: Concentrations between lakes are different.
## Testing mean annual concentrations

require(nlme)
fit <- lme(ResultMeasureValue ~ yr.fac, random = ~ 1 | MonitoringLocationIdentifier, 
            correlation=corAR1(form = ~ 1 | MonitoringLocationIdentifier),
            data = bmp, na.action = "na.omit")


#summary(fit)

# Mean seasonal concentrations per year in each lake
bmp3 <- aggregate(ResultMeasureValue ~ yr.fac + MonitoringLocationIdentifier, data = bmp, mean)

# Pawnee
fit.paw.aov <- aov(ResultMeasureValue ~ yr.fac, data = bmp.paw)
#summary(fit.paw.aov)
pander(fit.paw.aov)

res <- TukeyHSD(fit.paw.aov)
emphasize.strong.rows(which(res$yr.fac[,4] < 0.05 , arr.ind = TRUE))
pander(res)

fit.paw.lme <- lme(ResultMeasureValue ~ yr.fac, random = ~ 1 | mo,
            correlation=corAR1(form = ~ 1 | mo),
            data = bmp.paw, na.action = "na.omit")

anova(fit.paw.lme)
summary(fit.paw.lme)
```

\clearpage

```{r}
# Carter
fit.car.aov <- aov(data = bmp.car, ResultMeasureValue ~ yr.fac)
pander(fit.car.aov)

res <- TukeyHSD(fit.car.aov)
emphasize.strong.rows(which(res$yr.fac[,4] < 0.05 , arr.ind = TRUE))
pander(res)
```
\clearpage

```{r}
# Fremont
fit.fre.aov <- aov(data = bmp.fre, ResultMeasureValue ~ yr.fac)
pander(fit.fre.aov)

res <- TukeyHSD(fit.fre.aov)
emphasize.strong.rows(which(res$yr.fac[,4] < 0.05 , arr.ind = TRUE))
pander(res)


#summary(fit.fre)
#TukeyHSD(fit)

```

**Followup questions**
1. What happened in 2009 to decrease microcystin trends across all three lake systems?


\clearpage
### Q: What percentage of samples had detectable microcystins (> 0.15 mg/L for ELISA? What percentage of samples had total microcystin concentrations exceeding the WHO limit?

```{r DetMic}
## What percentage of samples have detectable microcystins
#For each location > for each year > for each month within a year

loc.list1 <- c()
loc.list2 <-c()
yr.list1 <- c() 
yr.list2 <- c() 
mo.list <- c() 
yr.perc.list <- c() 
mo.perc.list <- c()
yr.above.list <-c()
yr.total.list <-c()
mo.above.list <-c()
mo.total.list <-c()

for(i in unique(bmp$MonitoringLocationIdentifier)){
  tmp <- bmp[bmp$MonitoringLocationIdentifier == i,]
  
  
  for(j in unique(tmp$yr.fac)){
    yr.above <- dim(tmp[tmp$yr.fac == j & tmp$ResultMeasureValue >= 0.15,])[1]
    yr.total <- dim(tmp[tmp$yr.fac == j,])[1]
    yr.perc <- round((yr.above/yr.total)*100, digits = 2)
    

    yr.above.list <- append(yr.above.list, yr.above)
    yr.total.list <- append(yr.total.list, yr.total)
    
    yr.list1 <- append(yr.list1, rep(j,length(yr.perc)))
    yr.perc.list <- append(yr.perc.list, yr.perc)
    loc.list1 <- append(loc.list1, rep(i, length(yr.perc)))
    
    for(k in unique(tmp$mo)){
      tmp2 <- tmp[tmp$mo == k & tmp$yr.fac == j,]
      mo.above <- dim(tmp2[tmp2$ResultMeasureValue >= 0.15,])[1]
      mo.total <- dim(tmp2)[1]
      mo.perc <- round((mo.above/mo.total)*100, digits = 2)
      
      mo.above.list <- append(mo.above.list, mo.above)
      mo.total.list <- append(mo.total.list, mo.total)
      
      yr.list2 <- append(yr.list2, rep(j, length(mo.perc)))
      mo.list <- append(mo.list, rep(k, length(mo.perc)))
      mo.perc.list <- append(mo.perc.list, mo.perc)
      loc.list2 <- append(loc.list2, rep(i, length(mo.perc)))
    }
  }
}

yr.perc <- data.frame(loc.list1, yr.list1, yr.above.list, yr.total.list,yr.perc.list)
mo.perc <- data.frame(loc.list2, yr.list2, mo.list, 
                      mo.above.list, mo.total.list,mo.perc.list)
yr.ag1 <- aggregate(yr.above.list ~ loc.list1, data = yr.perc, sum)
yr.ag2 <- aggregate(yr.total.list ~ loc.list1, data = yr.perc, sum)

```

```{r WHOMic}

#For each location > for each year > for each month within a year

loc.list1 <- c()
loc.list2 <-c()
yr.list1 <- c() 
yr.list2 <- c() 
mo.list <- c() 
yr.perc.list <- c() 
mo.perc.list <- c()
yr.above.list <-c()
yr.total.list <-c()
mo.above.list <-c()
mo.total.list <-c()

for(i in unique(bmp$MonitoringLocationIdentifier)){
  tmp <- bmp[bmp$MonitoringLocationIdentifier == i,]
  
  
  for(j in unique(tmp$yr.fac)){
    yr.above <- dim(tmp[tmp$yr.fac == j & tmp$ResultMeasureValue >= 20,])[1]
    yr.total <- dim(tmp[tmp$yr.fac == j,])[1]
    yr.percent <- round((yr.above/yr.total)*100, digits = 2)
    

    yr.above.list <- append(yr.above.list, yr.above)
    yr.total.list <- append(yr.total.list, yr.total)
    
    yr.list1 <- append(yr.list1, rep(j,length(yr.percent)))
    yr.perc.list <- append(yr.perc.list, yr.percent)
    loc.list1 <- append(loc.list1, rep(i, length(yr.percent)))
    
    for(k in unique(tmp$mo)){
      tmp2 <- tmp[tmp$mo == k & tmp$yr.fac == j,]
      mo.above <- dim(tmp2[tmp2$ResultMeasureValue >= 20,])[1]
      mo.total <- dim(tmp2)[1]
      mo.percent <- round((mo.above/mo.total)*100, digits = 2)
      
      mo.above.list <- append(mo.above.list, mo.above)
      mo.total.list <- append(mo.total.list, mo.total)
      
      yr.list2 <- append(yr.list2, rep(j, length(mo.percent)))
      mo.list <- append(mo.list, rep(k, length(mo.percent)))
      mo.perc.list <- append(mo.perc.list, mo.percent)
      loc.list2 <- append(loc.list2, rep(i, length(mo.percent)))
    }
  }
}

yr.perc2 <- data.frame(loc.list1, yr.list1, 
                      yr.above.list, yr.total.list,
                      yr.perc.list)
mo.perc2 <- data.frame(loc.list2, yr.list2, mo.list, 
                      mo.above.list, mo.total.list,
                      mo.perc.list)

yr.perc2.ag1 <- aggregate(yr.above.list ~ loc.list1, data = yr.perc2, sum)
yr.perc2.ag2 <- aggregate(yr.total.list ~ loc.list1, data = yr.perc2, sum)

```

```{r EPAMic}

#For each location > for each year > for each month within a year

loc.list1 <- c()
loc.list2 <-c()
yr.list1 <- c() 
yr.list2 <- c() 
mo.list <- c() 
yr.perc.list <- c() 
mo.perc.list <- c()
yr.above.list <-c()
yr.total.list <-c()
mo.above.list <-c()
mo.total.list <-c()

for(i in unique(bmp$MonitoringLocationIdentifier)){
  tmp <- bmp[bmp$MonitoringLocationIdentifier == i,]
  
  
  for(j in unique(tmp$yr.fac)){
    yr.above <- dim(tmp[tmp$yr.fac == j & tmp$ResultMeasureValue >= 4,])[1]
    yr.total <- dim(tmp[tmp$yr.fac == j,])[1]
    yr.percent <- round((yr.above/yr.total)*100, digits = 2)
    

    yr.above.list <- append(yr.above.list, yr.above)
    yr.total.list <- append(yr.total.list, yr.total)
    
    yr.list1 <- append(yr.list1, rep(j,length(yr.percent)))
    yr.perc.list <- append(yr.perc.list, yr.percent)
    loc.list1 <- append(loc.list1, rep(i, length(yr.percent)))
    
    for(k in unique(tmp$mo)){
      tmp2 <- tmp[tmp$mo == k & tmp$yr.fac == j,]
      mo.above <- dim(tmp2[tmp2$ResultMeasureValue >= 4,])[1]
      mo.total <- dim(tmp2)[1]
      mo.percent <- round((mo.above/mo.total)*100, digits = 2)
      
      mo.above.list <- append(mo.above.list, mo.above)
      mo.total.list <- append(mo.total.list, mo.total)
      
      yr.list2 <- append(yr.list2, rep(j, length(mo.percent)))
      mo.list <- append(mo.list, rep(k, length(mo.percent)))
      mo.perc.list <- append(mo.perc.list, mo.percent)
      loc.list2 <- append(loc.list2, rep(i, length(mo.percent)))
    }
  }
}

yr.perc2 <- data.frame(loc.list1, yr.list1, 
                      yr.above.list, yr.total.list,
                      yr.perc.list)
mo.perc2 <- data.frame(loc.list2, yr.list2, mo.list, 
                      mo.above.list, mo.total.list,
                      mo.perc.list)

yr.perc2.ag1 <- aggregate(yr.above.list ~ loc.list1, data = yr.perc2, sum)
yr.perc2.ag2 <- aggregate(yr.total.list ~ loc.list1, data = yr.perc2, sum)

```


```{r DetMicFigure, fig.cap = "Annual percent detectable microcystins (> 0.15 ug/L; grey) and percent above proposed EPA limit (4 ug/L; red)."}

# Merge data frames for detected and above limit

dat2 <- merge(yr.perc,yr.perc2, by = c("loc.list1","yr.list1"))
dat2 <- dat2[,c(1,2,5,8)]
colnames(dat2) <- c("MonitoringLocationIdentifier", "Year", "DetMic", "WHOMic")

#m1 <- melt(dat2, ids = c("MonitoringLocationIdentifier", "Year"))
#head(m1)

p = ggplot(data = dat2, aes(Year, DetMic)) +
  scale_x_discrete(breaks = c(2005,2007,2009,2011,2013,2015)) + 
  geom_bar(stat = "identity") +
  geom_bar(data = dat2, aes(Year,WHOMic), 
           stat = "identity", col = "white",fill = "red") +
  facet_wrap(~MonitoringLocationIdentifier, ncol = 2) +
  ylab("Percent Detected")
print(p)

#ggsave(paste("./supporting-files/mic-trends.pdf"), plot = p, scale = 1,
       width = 7.5, height = 6, units = "in")

```


### Observations
Microcystin variants (> 0.15 ug/L) were detected in `r round((sum(yr.ag1[,2])/sum(yr.ag2[,2])*100), digits = 2)` of samples collected during the May-September sampling seasons in Pawnee, Fremont, and Carter Lakes between 2005-2015.

1. Between 2005 and 2015, microcystin variants were detected in `r round((sum(yr.ag1[3:4,2])/sum(yr.ag2[3:4,2])*100), digits = 2)`% of samples collected from Pawnee Lake during seasonal sampling between May and September (n = `r sum(yr.ag2[3:4,2])`). Of these, `r round((sum(yr.perc2.ag1[3:4,2])/sum(yr.perc2.ag2[3:4,2])*100), digits = 2)`% were greater than the WHO limit. 
2. Between 2005 and 2015, microcystin variants were detected in `r round((sum(yr.ag1[1:2,2])/sum(yr.ag2[1:2,2])*100), digits = 2)`% of samples collected from Fremont Lake during seasonal sampling between May and September (n = `r sum(yr.ag2[1:2,2])`). Of these, `r round((sum(yr.perc2.ag1[1:2,2])/sum(yr.perc2.ag2[1:2,2])*100), digits = 2)`% were greater than the WHO limit. 
3. Between 2005 and 2015, microcystin variants were detected in `r round((sum(yr.ag1[5,2])/sum(yr.ag2[5,2])*100), digits = 2)`% of samples collected from Carter Lake during seasonal sampling between May and September (n = `r sum(yr.ag2[5,2])`). Of these, `r round((sum(yr.perc2.ag1[5,2])/sum(yr.perc2.ag2[5,2])*100), digits = 2)`% were greater than the WHO limit.  


\clearpage

####Monthly trends between 2005-2009

```{r, fig.height = 8}
mos = c("05","06","07","08","09")

bmp = bmp[bmp$mo %in% mos,]

yrs = c("2006","2007","2008","2009")
bmp = bmp[bmp$yr.fac %in% yrs,]

bmp.paw <- bmp[grep("PAW", bmp$MonitoringLocationIdentifier),]
bmp.fre <- bmp[grep("FRM", bmp$MonitoringLocationIdentifier),]
bmp.car <- bmp[grep("CAR", bmp$MonitoringLocationIdentifier),]


seasonal.paw = ggplot(bmp.paw, aes(mo,ResultMeasureValue, 
                      shape = yr.fac, col = yr.fac)) +
  ylim(0,80) +
  geom_hline(yintercept = 4, col = "red", lty = 1, lwd = 1) +
  geom_jitter(cex = 1, col = "grey50", width = 0.2, height = 0.1)  +
  stat_summary(fun.data = "mean_cl_boot", color = "black", size = 1) +
  xlab("Month") + ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  annotate("text", y = Inf, x = Inf, label = "Pawnee Lake", vjust = 2, hjust = 1.25) +
  guides(shape = guide_legend(title = "Year"))

#print(seasonal.paw)

# PreAlum = black; postAlum = "grey"
seasonal.fre = ggplot(bmp.fre, aes(mo,ResultMeasureValue, 
                      shape = yr.fac, col = yr.fac)) +
  ylim(0,80) +
  #scale_x_continuous(breaks = c(2005:2015)) +
  geom_hline(yintercept = 4, col = "red", lty = 1, lwd = 1) +
  geom_jitter(cex = 1, col = "grey50", width = 0.2, height = 0.1)  +
  stat_summary(fun.data = "mean_cl_boot", color = "black", size = 1) +
  xlab("Month") + ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  annotate("text", y = Inf, x = Inf, label = "Fremont Lake No. 20", vjust = 2, hjust = 1.15) +
  guides(shape = guide_legend(title = "Year"))

#print(seasonal.fre)

seasonal.car = ggplot(bmp.car, aes(mo,ResultMeasureValue, 
                      shape = yr.fac, col = yr.fac)) +
  ylim(0,50) +
  #scale_x_continuous(breaks = c(05:09), labels = c("May", "Jun","Jul","Aug", "Sept"), guide = guide_legend(title = "Year")) +
  geom_hline(yintercept = 4, col = "red", lty = 1, lwd = 1) +
  geom_jitter(cex = 1, col = "grey50", width = 0.2, height = 0.1) +
  stat_summary(fun.data = "mean_cl_boot", color = "black", size = 1) +
  xlab("Month") + ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  annotate("text", y = Inf, x = Inf, label = "Carter Lake", vjust = 2, hjust = 1.25) +
  guides(shape = guide_legend(title = "Year"))

#print(seaonsal.car)

grid.arrange(seasonal.paw,seasonal.fre, seasonal.car, ncol = 1)

# statistics should control for variation between years and treatment

```

```{r, eval = FALSE}
grid.arrange(annual.paw,seasonal.paw,annual.fre,seasonal.fre,annual.car,seasonal.car, nrow = 3)

ggsave("./supporting-files/annualPaw4yr.pdf", annual.paw, 
       height = 2.5, width = 5, units = "in", scale = 1.5)
ggsave("./supporting-files/annualFre4yr.pdf", annual.fre, 
       height = 2.5, width = 5, units = "in", scale = 1.5)
ggsave("./supporting-files/annualCar4yr.pdf", annual.car, 
       height = 2.5, width = 5, units = "in", scale = 1.5)
ggsave("./supporting-files/seasonalPaw.pdf", seasonal.paw, 
       height = 2.5, width = 5, units = "in", scale = 1.5)
ggsave("./supporting-files/seasonalFre.pdf", seasonal.fre, 
       height = 2.5, width = 5, units = "in", scale = 1.5)
ggsave("./supporting-files/seasonalCar.pdf", seasonal.car, 
       height = 2.5, width = 5, units = "in", scale = 1.5)

```


\clearpage

```{r DetMicFigureMonth, fig.cap = "Monthly percent detectable microcystins (grey) and percent above WHO limit (red) between 2006-2009.", fig.height = 8}

# Merge data frames for detected and above limit

dat2 <- merge(mo.perc,mo.perc2, by = c("loc.list2","yr.list2","mo.list"))
dat2 <- dat2[,c(1,2,3,6,9)]
colnames(dat2) <- c("MonitoringLocationIdentifier", "Year", "Month", "DetMic", "WHOMic")

#m1 <- melt(dat2, ids = c("MonitoringLocationIdentifier", "Year"))
#head(m1)

p = ggplot(data = dat2, aes(Month, DetMic)) +
  geom_bar(stat = "identity") +
  geom_bar(data = dat2, aes(Month,WHOMic), 
           stat = "identity", col = "white",fill = "red") +
  facet_grid(Year~MonitoringLocationIdentifier) +
  ylab("Percent Detected")
print(p)

```

\clearpage

#### Weekly sampling between 2005-2009

```{r, fig.height = 8}

bmp.paw <- bmp.paw[bmp.paw$yr <=2007,]
bmp.fre <- bmp.fre[bmp.fre$yr <=2007,]
bmp.car <- bmp.car[bmp.car$yr <=2007,]

wkly.seasonal.paw = ggplot(bmp.paw, aes(wk,ResultMeasureValue)) +
  ylim(0,80) +
  #geom_hline(yintercept = 20, col = "red", lty = 1, lwd = 1) +
  stat_summary(fun.data = "mean_cl_boot", size = 1, geom = "line") +
  geom_point(cex = 1, col = "grey50", width = 0.2, height = 0.1)  +
  xlab("Week") + 
  ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  annotate("text", y = Inf, x = Inf, label = "Pawnee Lake", vjust = 2, hjust = 1.25)+
  guides(col = guide_legend(title = "Year"))

#print(seasonal.paw)

# PreAlum = black; postAlum = "grey"
wkly.seasonal.fre = ggplot(bmp.fre, aes(wk,ResultMeasureValue)) +
  ylim(0,80) +
  #geom_hline(yintercept = 20, col = "red", lty = 1, lwd = 1) +
  stat_summary(fun.data = "mean_cl_boot", size = 1, geom = "line") +
  geom_point(cex = 1, col = "grey50", width = 0.2, height = 0.1)  +
  xlab("Week") + 
  ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  annotate("text", y = Inf, x = Inf, label = "Fremont Lake No. 20", vjust = 2, hjust = 1.15)+
  guides(col = guide_legend(title = "Year"))

#print(seasonal.fre)

wkly.seasonal.car = ggplot(bmp.car, aes(wk,ResultMeasureValue)) +
  ylim(0,50) +
  #geom_hline(yintercept = 20, col = "red", lty = 1, lwd = 1) +
  stat_summary(fun.data = "mean_cl_boot", size = 1, geom = "line") +
  geom_point(cex = 1, col = "grey50", width = 0.2, height = 0.1)  +
  xlab("Week") + 
  ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  annotate("text", y = Inf, x = Inf, label = "Carter Lake", vjust = 2, hjust = 1.25)+
  guides(col = guide_legend(title = "Year"))

#print(seaonsal.car)

grid.arrange(wkly.seasonal.paw,wkly.seasonal.fre, wkly.seasonal.car, ncol = 1)

# statistics should control for variation between years and treatment

```

\clearpage

### Statistics
** Does the microcystin concentration change throughout the season?
```{r}
# use RMANOVA twith month nested within the year

fit.paw.lme <- lme(ResultMeasureValue ~ wk, random = ~ 1 | yr,
            correlation=corAR1(form = ~ 1 | yr),
            data = bmp.paw, na.action = "na.omit")
summary(fit.paw.lme)
anova(fit.paw.lme)

fit.fre.lme <- lme(ResultMeasureValue ~ wk, random = ~ 1 | yr,
            correlation=corAR1(form = ~ 1 | yr),
            data = bmp.fre, na.action = "na.omit")
summary(fit.fre.lme)
anova(fit.fre.lme)

fit.car.lme <- lme(ResultMeasureValue ~ wk, random = ~ 1 | yr,
            correlation=corAR1(form = ~ 1 | yr),
            data = bmp.car, na.action = "na.omit")
summary(fit.car.lme)
anova(fit.car.lme)

```

\clearpage

## LCMS data

```{r wsl-data-import}
wscmic <- read.csv("./data/wsc_toxindata.csv", header = T)
#str(wscmic)

wscmic[is.na(wscmic)] <-0

# Remove WSL QA/QC data
wscmic <- wscmic[-grep("LFM",wscmic$Sample_ID),]
wscmic <- wscmic[-grep("LD2",wscmic$Sample_ID),]

#remove samples without a location ID
wscmic <- wscmic[grep("21NEB001",wscmic$MonitoringLocationIdentifier),]

# subset data
wscmic <- wscmic[,c(2,4:14)]

#wscmic$Microcystin.total <- rowSums(wscmic[,8:12])
```

```{r crossreactivity}
#corrected = uncorrected LC/MS/MS concentration (ug/l)*ELISA cross-reactivity correction

wscmic$mcla <- wscmic$Microcystin.LA*.48
wscmic$mclf <- wscmic$Microcystin.LF*.72
wscmic$mclr <- wscmic$Microcystin.LR
wscmic$mclw <- wscmic$Microcystin.LW*1.02
wscmic$mcrr <- wscmic$Microcystin.RR*.53
wscmic$mc.total.cor <- rowSums(wscmic[,13:17])

wscmic$tmp <- as.Date(wscmic$Collection_Date, format = "%Y-%m-%d")
wscmic$yr <- as.factor(format(wscmic$tmp,'%Y'))                      
wscmic$mo <- as.factor(format(wscmic$tmp,'%m'))
#wscmic$jul.dat <- yday(wscmic$tmp)
wscmic$wk <- week(wscmic$tmp)


#write.csv(wscmic, "./supporting-files/20161019_wscmic-trimmed.csv", row.names = FALSE)
```

```{r}

# Convert time 
require(reshape2)
m <- melt(wscmic, ids = c("lake","Collection_Date"))


# subset data to only include BMP stations
sta <- c("21NEB001_WQX-LLP2PAWNEE03","21NEB001_WQX-LLP2PAWNEE05",
         "21NEB001_WQX-LLP1FRMT2004","21NEB001_WQX-LLP1FRMT2005",
         "21NEB001_WQX-LMT1CARTER03")

m <- m[m$MonitoringLocationIdentifier %in% sta,]

m$lake <- droplevels(m$lake)

m1 <- m[grep("mc.",m$variable),]
m1$jul.dat <- yday(m1$Collection_Date)

p1 = ggplot(m1, aes(jul.dat, value, col = variable, shape = yr)) +
  geom_hline(yintercept = 20, col = "red", lty = 1, lwd = 1) +
  geom_point(cex = 2) +
  scale_color_grey() +
  xlab("Julian date") + ylab("Concentration (ug/L)") +
  facet_wrap(~MonitoringLocationIdentifier, ncol = 2, scales = "free_y")

print(p1)


comp.paw <- m1[grep("PAWNEE",m1$MonitoringLocationIdentifier),]
comp.fre <- m1[grep("FRMT",m1$MonitoringLocationIdentifier),]
comp.car <- m1[grep("CAR",m1$MonitoringLocationIdentifier),]

p = ggplot(comp.paw[comp.paw$variable != "mc.total.cor",], 
       aes(jul.dat, value, fill = variable)) +
  geom_area(position = "stack") +
  xlim(min(m1$jul.dat),max(m1$jul.dat)) +
  xlab("Julian Date") + ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  facet_wrap(MonitoringLocationIdentifier~yr)

#print(p)

p1 <- ggplot(comp.fre[comp.fre$variable != "mc.total.cor",], 
       aes(jul.dat, value, fill = variable)) +
  geom_area(position = "stack") +
  xlim(min(m1$jul.dat),max(m1$jul.dat)) +
  xlab("Julian Date") + ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  facet_wrap(MonitoringLocationIdentifier~yr)

p2 <- ggplot(comp.car[comp.car$variable != "mc.total.cor",], 
       aes(jul.dat, value, fill = variable)) +
  geom_area(position = "stack") +
  xlim(min(m1$jul.dat),max(m1$jul.dat)) +
  xlab("Julian Date") + ylab(expression(paste("Microcystin Concentration (", mu,"g/L)"))) +
  facet_wrap(MonitoringLocationIdentifier~yr)

```


```{r}
p2 = ggplot(m1, aes(jul.dat, value, col = variable, shape = yr)) +
  geom_hline(yintercept = 20, col = "red", lty = 1, lwd = 1) +
  geom_point(cex = 2) +
  scale_color_grey() +
  geom_point(data = m1[m1$variable == "Microcystin.total",],
             aes(jul.dat,value, shape = yr), pch = 20, col = "red", cex = 3) +
  xlab("Julian date") + ylab(expression(paste("Concentration (",mu,"g/L)"))) +
  ylim(0,20)+
  facet_wrap(~lake, ncol = 2, scales = "free_y")


#p2$data$lake <- factor(p2$data$lake, levels = c("PAWNEE WEST","PAWNEE EAST","FREMONT WEST","FREMONT EAST","CARTER"))

#print(p2)

#grid.arrange(p1,p2)
```

## Method comparision

```{r, fig.height=4,fig.width=4,fig.cap="Relationship between total microcystin concentration as measured by LC/MS/MS and ELISA (n = 110). Total microcystin concentrations measured via LC/MS/MS were calculated by summing the cross-reactivity corrected concentration of each microcystin congener."}

dat <- dat[dat$CharacteristicName == "Microcystin",]

mic.merge <- merge(dat,wscmic, 
                   by.x = c("MonitoringLocationIdentifier","tmp"), 
                   by.y = c("MonitoringLocationIdentifier","Collection_Date"))

#mic.merge2 <- mic.merge2[,c(1,2,9,35,36,37,38,67,68,69,70,72,73,74,80,81,82,83,84,85,86)]

# with log scaling, remove data below detection limit
p1 = ggplot(mic.merge[mic.merge$ResultMeasureValue >=0.15,], aes(x = mc.total.cor, y = ResultMeasureValue)) +
 annotation_logticks(base = 10, sides = "bl", scaled = TRUE, 
                      short = unit(0.1, "cm"), 
                      mid = unit(0.2, "cm"), 
                      long = unit(0.3, "cm"), 
                      colour = "black", size = 0.5, 
                      linetype = 1, alpha = 1, color = "grey25") +
  scale_y_log10(breaks = c(0.01,0.1,1,10,100), labels = c(0.01,0.1,1,10,100)) +
  scale_x_log10(breaks = c(0.01,0.1,1,10,100), labels = c(0.01,0.1,1,10,100)) +
  geom_abline(intercept = 0, slope = 1, col = "black", lwd = 1, lty = 1) +
  geom_point(cex = 3) +
  xlab(expression(paste("Log10 Total Corrected Microcystin by LC - MS/MS (", mu,"g/L)"))) +
  ylab(expression(paste("Log10 Total Microcystin by ELISA (" , mu,"g/L)"))) +
  annotate("text",x = 145, y = 100,label = "1:1", col = "black") +
  annotate("text", x = Inf, y = Inf, label = "p = 0.011", vjust = 2, hjust = 1.25)

print(p1)

#ggsave("./supporting-files/FigS2Correlation.pdf", p1, units = "in", height = 3, width = 3, scale = 2)

#length(mic.merge$mc.total.cor[mic.merge$mc.total.cor >0])/length(mic.merge$mc.total.cor)
#Only 20% (19.8) of samples had comparible microcystin concencentrations; 80% of those samples contained microcystins via ELISA suggesting microcystin congeners other than the most common five variants in the samples

p2 = ggplot(mic.merge[mic.merge$mc.total.cor >0,], aes(x = mc.total.cor, y = ResultMeasureValue, col = lake, shape = lake)) +
  scale_x_log10()+
  scale_y_log10() +
  geom_abline(intercept = 0, slope = 1, col = "red", lwd = 1, lty = 1) +
  geom_point(cex = 2) +
  xlab("Total Corrected Microcystin by LC/MS/MS (ug/L)") +
  ylab("Total Microcystin by ELISA (ug/L)")

#print(p2)

#cor(mic.merge$mc.total.cor, mic.merge$mic.con, method = "spearman")
#install.packages("psych"");corr.test()
mc.merge.stats <- mic.merge[mic.merge$ResultMeasureValue >=0.15,]
cortest <- cor.test(mc.merge.stats$mc.total.cor, mc.merge.stats$ResultMeasureValue, method = "pearson", conf.level = 0.95)
cortest
```

LCMS and ELISA are significantly correlated (t = `r cortest[[1]]`,df = `r cortest[[2]]` ,p = `r cortest[[3]]`) following cross-reactivity correction (Loftin et al 2008).

```{r, include = FALSE}
# This is the old data set

ndeqmic <- read.csv("./data/ndeq_toxindata.csv", header = T)
#str(ndeqmic)


ndeqmic$tmp <- as.Date(ndeqmic$sample.date, format = "%Y-%m-%d")
ndeqmic$yr <- as.factor(format(ndeqmic$tmp,'%Y'))                      
ndeqmic$jul.dat <- yday(ndeqmic$tmp)

ndeqmic <- ndeqmic[-grep(">", ndeqmic$mic.con),]
ndeqmic$mic.con <- as.numeric(as.character(ndeqmic$mic.con))

p = ggplot(ndeqmic, aes(jul.dat,mic.con, col = yr)) +
  geom_hline(yintercept = 20, col = "red", lty = 1, lwd = 1) +
  geom_point(cex = 2) +
  scale_color_grey() +
  xlab("Julian date") + ylab("Microcystin Concentration (ug/l)") +
  facet_wrap(~lake, ncol = 2)
#p$data$lake <- factor(p$data$lake, levels = c("PAWNEE WEST","PAWNEE EAST","FREMONT WEST", "FREMONT EAST","CARTER"))

#print(p)

```

```{r, fig.height=4,fig.width=4,fig.cap="Relationship between total microcystin concentration as measured by LC/MS/MS and ELISA (n = 110). Total microcystin concentrations measured via LC/MS/MS were calculated by summing the cross-reactivity corrected concentration of each microcystin congener.", include = FALSE}
#Merge data from NDEQ and WSC
mic.merge <- merge(ndeqmic,wscmic, 
                  by.x = c("lake","sample.date"), 
                   by.y = c("lake","Collection_Date"), all = T)

p1 = ggplot(mic.merge, aes(x = mc.total.cor, y = mic.con)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(intercept = 0, slope = 1, col = "black", lwd = 1, lty = 1) +
  geom_point(cex = 3) +
  xlab(expression(paste("Total Corrected Microcystin by LC - MS/MS (", mu,"g/L)"))) +
  ylab(expression(paste("Total Microcystin by ELISA (" , mu,"g/L)"))) +
  annotate("text",x = 145, y = 100,label = "1:1", col = "black") +
  annotate("text", x = 0.01, y = 100, label = "Stats")

print(p1)

ggsave("./supporting-files/FigS2Correlation.pdf", p1, units = "in", height = 3, width = 3, scale = 2)

p2 = ggplot(mic.merge[mic.merge$mc.total.cor >0,], aes(x = mc.total.cor, y = mic.con, col = lake, shape = lake)) +
  scale_x_log10()+
  scale_y_log10() +
  geom_abline(intercept = 0, slope = 1, col = "red", lwd = 1, lty = 1) +
  geom_point(cex = 2) +
  xlab("Total Corrected Microcystin by LC/MS/MS (ug/L)") +
  ylab("Total Microcystin by ELISA (ug/L)")

#print(p2)

#cor(mic.merge$mc.total.cor, mic.merge$mic.con, method = "spearman")
cortest <- cor.test(mic.merge$mc.total.cor, mic.merge$mic.con, method = "spearman", conf.level = 0.95)
```

```{r, eval = FALSE, include = FALSE}

#length(mic.merge$mc.total.cor[mic.merge$mc.total.cor <0.1])
#length(mic.merge$mc.total.cor)


int <- mic.merge[mic.merge$ResultMeasureValue < 10 & mic.merge$mc.total.cor > 25,c("lake","ActivityStartDate","ResultMeasureValue","mc.total.cor","Anatoxin.A","Cylindrospermopsin","Saxitoxin")]

int <- mic.merge[mic.merge$mc.total.cor > 0,c("lake","ActivityStartDate","ResultMeasureValue","mc.total.cor","Anatoxin.A","Cylindrospermopsin","Saxitoxin")]

#int <- cbind(int[,1:3],int[,12],int[,4:6])
colnames(int) <- c("Lake","Sample Date","ELISA MIC","LC/MS/MS MIC","ATX", "CYL", "SAX")

row.names(int) <- NULL
int <- na.omit(int)
pander(int)
```
\newpage

## Other algal toxins
```{r}
toxins <- c("Anatoxin.A", "Cylindrospermopsin","Saxtoxin")

m2 <- m[toxins %in% m$variable]

lvls <- read.csv("./data/state_regs.csv", header = T)
lvls <- lvls[3:5,3:4]
lvls$toxin <- c("Anatoxin.A","Cylindrospermopsin","Saxitoxin")

p1 = ggplot(m2[m2$value >0,], aes(mo, value, shape = yr)) +
  #geom_hline(data = lvls, yintercept = lvls$PHA, lwd = 1)+
  #geom_hline(data = lvls, yintercept = lvls$NCA, lwd = 2)+
  geom_point(cex = 2) +
  scale_color_discrete(c("red","blue","green"))+
  xlab("Month") + ylab(expression(paste("Cyanotoxin Concentration (", mu,"g/L)"))) +
  facet_grid(~variable, scales = "free_y") 

#print(p1)

p2 = ggplot(m2[m2$value >0,], aes(jul.dat, value, col = yr, shape = lake)) +
  geom_point(cex = 2) +
  scale_color_discrete(c("red","blue","green")) +
  xlab("Julian date") + ylab("Concentration (ug/L)") +
  ylim(0,15)+
  facet_grid(~variable, scales = "free_y")

#print(p2)


p3 = ggplot(m2[m2$value > 0,], aes(jul.dat, value, col = yr, shape = lake)) +
  geom_point(cex = 2) +
  scale_color_discrete(c("red","blue","green")) +
  xlab("Julian date") + ylab("Concentration (ug/L)") +
  ylim(0,1)+
  facet_grid(~variable, scales = "free_y")

#print(p3)

#grid.arrange(p1,p2,p3)

```

1. The State of Nebraska only monitors the concentration of microcystins as an aggregate group and issues health advisories for recreational water bodies as a total toxin concentrations greater than 20 ug/L. 
Other states issue both public health advisories (PHA) and no contact advisories (NCA) depending on the type of cyanobacterial toxin. 
For example, Ohio regularly tracks microcystin-LR, anatoxin-a, saxitoxin, and cylindrospermopsin and issues both PHA and NCA throughout the season (Table 1, next page).
2. **Anatoxin a**: 
The reporting limit for our method is; 
`r length(m2$value[m2$variable == "Anatoxin.A" & m2$value >= 0.01])` samples (~ `r round(length(m2$value[m2$variable == "Anatoxin.A" & m2$value >= 0.01])/length(m2$value[m2$variable == "Anatoxin.A"]) * 100, digits = 1)` %) contained an a reportable anatoxin concentration (mean =`r round(mean(m2$value[m2$variable == "Anatoxin.A"]), digits = 1)` ug/L). 
The OH PHA advisory for anatoxin a is 80 ug/L; 
In this study, `r length(m2$value[m2$variable == "Anatoxin.A" & m2$value >= 80])` samples (~ `r round(length(m2$value[m2$variable == "Anatoxin.A" & m2$value >= 80])/length(m2$value[m2$variable == "Anatoxin.A"]) * 100, digits = 1)` %) contained an anatoxin concentration greater than the PHA advisory (mean =`r round(mean(m2$value[m2$variable == "Anatoxin.A"]), digits = 1)` ug/L). 
The OH NCA advisory is 300 ug/L; 
In these samples, `r length(m2$value[m2$variable == "Anatoxin.A" & m2$value >=300])` samples (~ `r round(length(m2$value[m2$variable == "Anatoxin.A" & m2$value >= 300])/length(m2$value[m2$variable == "Anatoxin.A"]) * 100, digits = 1)` %) contained an anatoxin concentration greater than the NCA advisory.
3. **Cylindrospermopsin**: The OH PHA advisory for cylindrospermopsin is 5 ug/L; 
In this study, `r length(m2$value[m2$variable == "Cylindrospermopsin" & m2$value >=5])` samples (~ `r round(length(m2$value[m2$variable == "Cylindrospermopsin" & m2$value >= 5])/length(m2$value[m2$variable == "Cylindrospermopsin"]) * 100, digits =1)` %) contained a cylindrospermopsin concentration greater than the PHA advisory (mean = `r round(mean(m2$value[m2$variable == "Cylindrospermopsin"]),digits = 1)` ug/L). 
The OH NCA advisory for cylindrospermopsin is 20 ug/L; 
In these samples, `r length(m2$value[m2$variable == "Cylindrospermopsin" & m2$value >=20])` samples (~ `r round(length(m2$value[m2$variable == "Cylindrospermopsin" & m2$value >= 20])/length(m2$value[m2$variable == "Cylindrospermopsin"]) * 100, digits = 1)` %) contained a cylindrospermopsin concentration greater than the NCA advisory.
4. **Saxitoxin**: The OH PHA advisory for saxitoin is 0.8 ug/L; 
In this study, `r length(m2$value[m2$variable == "Saxitoxin" & m2$value >=0.8])` samples (~ `r round(length(m2$value[m2$variable == "Saxitoxin" & m2$value >= 0.8])/length(m2$value[m2$variable == "Saxitoxin"]) * 100, digits = 1)` %) contained a saxitoxin concentration greater than the PHA advisory (mean = `r round(mean(m2$value[m2$variable == "Saxitoxin"]), digits = 1)` ug/L). 
The OH NCA advisory for saxitoin is 3 ug/L; 
In these samples, `r length(m2$value[m2$variable == "Saxitoxin" & m2$value >=3])` samples (~ `r round(length(m2$value[m2$variable == "Saxitoxin" & m2$value >= 3])/length(m2$value[m2$variable == "Saxitoxin"]) * 100, digits = 1)` %) contained a saxitoxin concentration greater than the NCA advisory.

\newpage
```{r}
regs <- read.csv("./data/state_regs.csv", header = TRUE)

require(pander)
set.caption("Selected state regulations for cyanotoxins.PHA = Personal Health Advisory; NCA = No Contact Advisory")
pander(regs)
```




# Results: Climatic and environmental data by lake system

## Climatic, environmental, and biological data sets
*Are there environmental parameters responsible for microcystin production?*
```{r}
# Read in data table

# Make pander table

# Save table
```


```{r LoadEnvData}
### ENVIRONMENTAL ###
# 'dat' data file contains all the environmental information needed for the analysis
dat <- read.csv("./data/result.csv", header = TRUE)

# Do some data cleanup for this analysis

dat$ResultMeasureValue <- as.numeric(as.character(dat$ResultMeasureValue))

## Convert ActivityStartDate to a Juilan value to look at annual and decadal patterns
require(lubridate)
require(ggplot2)

dat$tmp <- as.Date(dat$ActivityStartDate, format = "%Y-%m-%d") #Converts data to date structure
dat$yr <- as.numeric(format(dat$tmp,'%Y'))         #Time is 
dat$yr.fac <- as.factor(as.numeric(format(dat$tmp,'%Y')))         #Time is now numeric
dat$mo <- as.factor(format(dat$tmp,'%m'))
dat$jul.dat <- yday(dat$tmp)                                  #Time is now in Julian
dat$wk <- week(dat$tmp)

```

```{r LoadClimData}
### CLIMATIC ###
noaa.dat <- read.csv("./data/noaa_data.csv", header = TRUE)

## Add time dimensions to the data set
noaa.dat$DATE <- as.factor(noaa.dat$DATE)
noaa.dat$tmp <- as.Date(noaa.dat$DATE, format = "%Y%m%d")
noaa.dat$yr <- as.numeric(format(noaa.dat$tmp,'%Y'))  
noaa.dat$yr.fac <- as.factor(as.numeric(format(noaa.dat$tmp,'%Y')))
noaa.dat$mo <- as.factor(format(noaa.dat$tmp,'%m'))
noaa.dat$wk <- week(noaa.dat$tmp)
noaa.dat$jul.dat <- yday(noaa.dat$tmp)

## Trim up data set to include only data through 2015
noaa.dat <- noaa.dat[noaa.dat$yr <=2015 & noaa.dat$yr >=2005,]
#noaa.dat <- noaa.dat[-grep("EMERALD",noaa.dat$STATION_NAME),]
```

\newpage

## Pawnee Lake
```{r PawEnv}
paw.dat <- dat[grep("PAW",dat$MonitoringLocationIdentifier),]
paw.dat$MonitoringLake <- "PAWNEE"

#unique(paw.dat$OrganizationIdentifier)
#unique(paw.dat$yr)

## Focus on data collected between 2005 to 2015
paw.dat <- paw.dat[paw.dat$yr >= 2005 & paw.dat$yr < 2010,]

## Data now needs to be reshaped to merge environmental and climate data

## Pull out appropriate columns
vars <- c("tmp", "yr.fac","mo", "wk", "CharacteristicName", "ResultMeasureValue", "ActivityDepthHeightMeasure.MeasureValue","OrganizationIdentifier","ProjectIdentifier","MonitoringLocationIdentifier")

## Select Environmental Parameters of interest

env.parm <- c("Microcystin" ,"Alkalinity, total","Aluminum","Ammonia-nitrogen","Chlorophyll a",
          "Chlorophyll a (probe)","Depth","Depth, Secchi disk depth","Dissolved oxygen (DO)","Inorganic nitrogen (nitrate and nitrite)",
          "pH","Phosphorus","Orthophosphate","Specific conductance","Temperature, water",
          "Total suspended solids", "Turbidity","Kjeldahl nitrogen")

# Subset the data frames for surface (< 0.1m) and depth profiles
paw.dat2 <- paw.dat[,vars]
paw.dat2$wk.fac <- as.factor(paw.dat2$wk)
paw.dat2$CharacteristicName <- as.character(paw.dat2$CharacteristicName)

paw.dat2 <- paw.dat2[paw.dat2$CharacteristicName %in% env.parm,]
paw.dat2$ActivityDepthHeightMeasure.MeasureValue[is.na(paw.dat2$ActivityDepthHeightMeasure.MeasureValue)] <- 0

```

```{r PawWaterColumn}
#unique(paw.dat2$MonitoringLocationIdentifier); unique(paw.dat2$CharacteristicName)

paw.dat2$jul.dat <- yday(paw.dat2$tmp)

parms.exclude = c("Depth, Secchi disk depth", "Depth","Aluminum", "Microcystin", "Chlorophyll a")
parms.test = c("Chlorophyll a (probe)","Dissolved oxygen (DO)", "Temperature, water","Kjeldahl nitrogen")
paw.dat.wtc <- paw.dat2[paw.dat2$CharacteristicName %in% parms.test,]

p <- ggplot(data = paw.dat2, 
            aes(x = tmp, y = ActivityDepthHeightMeasure.MeasureValue)) +
  geom_point() +
  scale_y_reverse() + 
  ylab("Lake Depth (m)") +
  xlab("Date") +
  facet_grid(CharacteristicName~MonitoringLocationIdentifier, scales = "free_y")

p$data$MonitoringLocationIdentifier <- factor(p$data$MonitoringLocationIdentifier, levels = c("COEOMAHA_WQX-PAWLKND1","COEOMAHA_WQX-PAWLKML1","COEOMAHA_WQX-PAWLKUP1"))

print(p)




p1 = ggplot(data = paw.dat.wtc[paw.dat.wtc$CharacteristicName == "Temperature, water",],
       aes(x = jul.dat, y = ActivityDepthHeightMeasure.MeasureValue)) +
  geom_point() +
  scale_y_reverse() + 
  ylab("Lake Depth (m)") +
  xlab("Date") +
  facet_grid(yr.fac~MonitoringLocationIdentifier, scales = "free_y")

p1$data$MonitoringLocationIdentifier <- factor(p1$data$MonitoringLocationIdentifier, levels = c("COEOMAHA_WQX-PAWLKND1","COEOMAHA_WQX-PAWLKML1","COEOMAHA_WQX-PAWLKUP1"))

print(p1)

p2 = ggplot(data = paw.dat.wtc[paw.dat.wtc$CharacteristicName == "Temperature, water" & paw.dat.wtc$ActivityDepthHeightMeasure.MeasureValue<0.5,],
       aes(x = wk, y = ResultMeasureValue)) +
  geom_point() +
  ylab("Water Temperature (Deg C)") +
  xlab("Date") 
#+
 # facet_grid(yr.fac~MonitoringLocationIdentifier, scales = "free_y")

p2$data$MonitoringLocationIdentifier <- factor(p1$data$MonitoringLocationIdentifier, levels = c("COEOMAHA_WQX-PAWLKND1","COEOMAHA_WQX-PAWLKML1","COEOMAHA_WQX-PAWLKUP1"))

print(p2)



```

\newpage

```{r}
# Print a new plot for each variable's depth profile...

mic = ggplot(data = paw.dat2[paw.dat2$CharacteristicName == "Microcystin" &paw.dat2$OrganizationIdentifier == "21NEB001_WQX" ,],
       aes(x = jul.dat, y = ResultMeasureValue)) +
  geom_path() +
  geom_point(color="white", size = 1) +
  geom_point(color = "black", size = 0.8)+
  ylab("Total Microcystin (ug/L)") +
  xlab("Date") +
  facet_grid(yr.fac~MonitoringLocationIdentifier, scales = "free_y")

print(mic)

```

```{r PawEpi}
# Evaluate just the data in the epilimnion
paw.epi <- paw.dat2[paw.dat2$ActivityDepthHeightMeasure.MeasureValue <= 0.5 ,]
paw.epi$yr.fac <- factor(paw.epi$yr.fac)
paw.epi$mo <- factor(paw.epi$mo)


# a first look at parameters within the epilimnion

p <- ggplot(data = paw.epi, aes(tmp, ResultMeasureValue)) +
  geom_point() +
  facet_wrap(~ CharacteristicName, scales = "free_y")
print(p)


## To merge data sets, first separate by projects then remerge
#paw.dat.ndeq <- paw.dat2[paw.dat2$OrganizationIdentifier != "COEOMAHA_WQX",]
#paw.dat.ndeq <- paw.dat.ndeq[,c(1:4,6,10)]
#colnames(paw.dat.ndeq) <- c("tmp","yr.fac","mo","wk","Microcystin","wk.fac")

paw.dat.coe <- paw.epi
## COE data only 2005-2009
paw.dat.coe <- paw.dat.coe[,-1]

require(reshape)
paw.dat.coe.reshape <- melt(paw.dat.coe)
paw.dat.coe.reshape <- paw.dat.coe.reshape[paw.dat.coe.reshape$variable == "ResultMeasureValue",]
#paw.dat.coe.reshape <- na.omit(paw.dat.coe.reshape)

# Evaluate the dataset for missing values using length of dataset
paw.dat.coe.reshape2 <- cast(data = paw.dat.coe.reshape, yr.fac + wk.fac ~ CharacteristicName)

# use cast function to add values to datafile
paw.dat.coe.reshape2 <- cast(data = paw.dat.coe.reshape, yr.fac + wk.fac ~ CharacteristicName, fill = "NA", add.missing = TRUE, fun.aggregate = mean)

#paw.dat.coe.reshape2 <- dcast(data = paw.dat.coe.reshape, yr.fac + wk.fac ~ CharacteristicName, value.var = paw.dat.coe.reshape$value, fill = "NA", drop = TRUE, fun.aggregate = mean)


# Final Pawnee environmental dataset
paw.dat.final <- paw.dat.coe.reshape2
head(paw.dat.final)
str(paw.dat.final)

# Function to convert factors to numeric data
as.numeric.factor <- function(x) {as.numeric(levels(x))[x]}

yr <- as.numeric.factor(paw.dat.final$yr.fac)
wk <- as.numeric.factor(paw.dat.final$wk.fac)

paw.dat.final <- cbind(yr,wk,sapply(paw.dat.final[,3:19],as.numeric.factor))

# Can also evalutate correlation between parameters
#install.packages("corrplot)")
#require(corrplot)
#cor1 <- cor(paw.dat.final[,3:ncol(paw.dat.final)])

```

**Environmental data summary**
The Army Corp of Enginners (Omaha, NE) collected data on the Pawnee Lake Reservoir between 2005 and 2009. Of the `r length(unique(paw.dat$CharacteristicName))-1` parameters available, we selected `r length(env.parm)-1` collected at the surface (< 0.1 m) for our analysis of microcystin trends (NDEQ) in Pawnee Lake.

```{r EnvParmTable}
require(pander)
EnvParmTable <- data.frame(paw.dat$OrganizationIdentifier[paw.dat$CharacteristicName %in% env.parm],
                           paw.dat$CharacteristicName[paw.dat$CharacteristicName %in% env.parm],
                           paw.dat$ResultMeasure.MeasureUnitCode[paw.dat$CharacteristicName %in% env.parm],
                           paw.dat$MethodDescriptionText[paw.dat$CharacteristicName %in% env.parm]
                           ,row.names = NULL)

EnvParmTable <- EnvParmTable[!duplicated(EnvParmTable),]
#EnvParmTable <- EnvParmTable[-c(1,12),]
row.names(EnvParmTable) <- NULL
colnames(EnvParmTable) <- c("Organization","Characteristic","Unit of Measure","Method URL")

set.caption("Selected variables for analysis")
set.alignment("left")
pander(EnvParmTable)
```

```{r PawNoaa}
## Select local data from 2005-2009
noaa.dat.paw <- noaa.dat[grep("LINCOLN",noaa.dat$STATION_NAME),]
noaa.dat.paw <- noaa.dat.paw[noaa.dat.paw$yr >= 2005 & noaa.dat.paw$yr <= 2009,]

noaa.dat.paw <- noaa.dat.paw[,c(7,17,22,32,37,42,212:217)]
noaa.dat.paw$mo.fac <- as.factor(noaa.dat.paw$mo)
noaa.dat.paw$wk.fac <- as.factor(noaa.dat.paw$wk)
noaa.dat.paw.reshape <- melt(noaa.dat.paw, ids=c("yr.fac","mo.fac","wk.fac"))

envs <- c("PRCP","SNOW","TSUN","TMAX","TMIN","WESD")
noaa.dat.paw.reshape <- noaa.dat.paw.reshape[noaa.dat.paw.reshape$variable %in% envs,]

noaa.dat.paw.reshape$value[noaa.dat.paw.reshape$value == -9999] <- NA
#noaa.dat.paw.reshape <- na.omit(noaa.dat.paw.reshape)

require(reshape)
# calculate wkly average precipitation, cloud cover
noaa.dat.paw <- cast(noaa.dat.paw.reshape, yr.fac + wk.fac ~ variable, mean)
head(noaa.dat.paw)

```

**NOAA Climatic data**
```{r NOAATable}
NOAATable <- data.frame(c("Preciptation","Snowfall","Time with sun","Max Temperature","Min Temperature","Water Equivalent Snow"), colnames(noaa.dat.paw)[3:8],c("in","in","minutes","deg C", "deg C", "in"))
colnames(NOAATable) <-c("Measure", "Abbrevation","Unit")

set.caption("Climatic variables")
set.alignment('left')
pander(NOAATable)
```

```{r, SamplingMap}
# Set spatial boundaries

# Read bathymetry data

# Make lake plot

# Plot sampling points

# Save map

```


\clearpage


```{r PawMerge}
# Merge data sets by year and week to have complete multivariate set
paw.dat.merged <- merge(paw.dat.final, noaa.dat.paw, by.x = c("yr","wk"), by.y = c("yr.fac","wk.fac"))

#Check the structure of the data set
str(paw.dat.merged)

# merge the LCMS data with the above dataset
paw.lcms <- mic.merge[grep("PAW",mic.merge$lake),]
paw.lcms <- paw.lcms[,c("lake","yr.y","wk","mcla","mclf","mclr","mclw","mcrr")]
paw.lcms[,"yr.y"] <- as.numeric.factor(paw.lcms[,"yr.y"]) 

# remember there are 2 data points per sample for the east and west beach
paw.dat.merged2 <- merge(paw.dat.merged,paw.lcms, by.x = c("yr","wk"), by.y = c("yr.y","wk"), all = T)

#Subset to the 2006-2009 years
paw.dat.merged3 <- paw.dat.merged2[paw.dat.merged$yr != "2005",]
paw.dat.merged3 <- paw.dat.merged3[,-22]

```

```{r PawMultivariateStats}
# For the initial analysis, remove microcystin composition
paw.dat4 <- paw.dat.merged3[,1:(ncol(paw.dat.merged3)-5)]
paw.dat5 <- melt(paw.dat4, id = c("yr","wk","Microcystin"))
paw.dat5 <- na.omit(paw.dat5)
ggplot(data = paw.dat5, aes(x = value, y = Microcystin, col = yr)) +
  geom_point()+
  facet_wrap(~variable, scales = "free_x")

paw.dat5 <- melt(paw.dat4, id = c("yr","wk"))
paw.dat5 <- na.omit(paw.dat5)
paw.dat5$value <- as.numeric(paw.dat5$value)
ggplot(data = paw.dat5, aes(x = yr, y = value, col = wk)) +
  stat_summary(fun.data = "mean_cl_boot")+
  facet_wrap(~variable, scales = "free_y")
```
\newpage

## Fremont Lake
```{r FreEnv}
fre.dat <- dat[grep("FRMT",dat$MonitoringLocationIdentifier),]
fre.dat$MonitoringLake <- "FREMONT"

## Substantial amount of duplication in dataset, needs to be removed and/or addressed. Maybe at different sampling depths?

#unique(fre.dat$OrganizationIdentifier)
#unique(fre.dat$yr)

## Focus on data collected between 2005 to 2015
fre.dat <- fre.dat[fre.dat$yr >= 2005 & fre.dat$yr < 2010,]

## Data now needs to be reshaped to merge environmental and climate data

## Pull out appropriate columns
vars <- c("tmp", "yr.fac","mo", "wk", "CharacteristicName", "ResultMeasureValue","ResultMeasure.MeasureUnitCode", "ActivityDepthHeightMeasure.MeasureValue","OrganizationIdentifier","ProjectIdentifier","MonitoringLocationIdentifier")

## Select Environmental Parameters of interest; this list will be different than Pawnee

env.parm <- c("Microcystin" ,"Alkalinity","Aluminum", "Ammonia-nitrogen","Chlorophyll a",
          "Depth, bottom","Depth, Secchi disk depth","Dissolved oxygen (DO)","Inorganic nitrogen (nitrate and nitrite)","Nitrate & Nitrite",
          "pH","Phosphate-phosphorus","Orthophosphate","Specific conductance","Temperature, water",
          "Total suspended solids", "Turbidity","Kjeldahl nitrogen", "Total Nitrogen","Total Phosphorus")

# Subset the data frames for surface (< 0.1m) and depth profiles
fre.dat2 <- fre.dat[,vars]
fre.dat2$wk.fac <- as.factor(fre.dat2$wk)
fre.dat2$CharacteristicName <- as.character(fre.dat2$CharacteristicName)

fre.dat2 <- fre.dat2[fre.dat2$CharacteristicName %in% env.parm,]
fre.dat2$ActivityDepthHeightMeasure.MeasureValue[is.na(fre.dat2$ActivityDepthHeightMeasure.MeasureValue)] <- 0
fre.epi <- fre.dat2[fre.dat2$ActivityDepthHeightMeasure.MeasureValue <= 0.1 ,]
fre.epi$yr.fac <- factor(fre.epi$yr.fac)
fre.epi$mo <- factor(fre.epi$mo)

fre.epi <- na.omit(fre.epi)

# a first look at parameters

p <- ggplot(data = fre.epi, aes(tmp, ResultMeasureValue)) +
  geom_point() +
  facet_wrap(~ CharacteristicName, scales = "free_y")
print(p)


## To merge data sets, first separate by projects then remerge
#fre.dat.ndeq <- fre.dat2[fre.dat2$OrganizationIdentifier != "COEOMAHA_WQX",]
#fre.dat.ndeq <- fre.dat.ndeq[,c(1:4,6,10)]
#colnames(fre.dat.ndeq) <- c("tmp","yr.fac","mo","wk","Microcystin","wk.fac")

fre.dat.coe <- fre.epi
## COE data only 2005-2009
fre.dat.coe <- fre.dat.coe[,-1]

require(reshape)
fre.dat.coe.reshape <- melt(fre.dat.coe)
fre.dat.coe.reshape <- fre.dat.coe.reshape[fre.dat.coe.reshape$variable == "ResultMeasureValue",]
#fre.dat.coe.reshape <- na.omit(fre.dat.coe.reshape)

# Evaluate the dataset for missing values using length of dataset
fre.dat.coe.reshape2 <- cast(data = fre.dat.coe.reshape, yr.fac + wk.fac ~ CharacteristicName)

# use cast function to add values to datafile
fre.dat.coe.reshape2 <- cast(data = fre.dat.coe.reshape, yr.fac + wk.fac ~ CharacteristicName, fill = "NA", add.missing = TRUE, fun.aggregate = mean)

#fre.dat.coe.reshape2 <- dcast(data = fre.dat.coe.reshape, yr.fac + wk.fac ~ CharacteristicName, value.var = fre.dat.coe.reshape$value, fill = "NA", drop = TRUE, fun.aggregate = mean)


# Final Fremont environmental dataset
fre.dat.final <- fre.dat.coe.reshape2
head(fre.dat.final)
str(fre.dat.final)

# Function to convert factors to numeric data
as.numeric.factor <- function(x) {as.numeric(levels(x))[x]}

yr <- as.numeric.factor(fre.dat.final$yr.fac)
wk <- as.numeric.factor(fre.dat.final$wk.fac)

fre.dat.final <- cbind(yr,wk,sapply(fre.dat.final[,3:19],as.numeric.factor))

# Can also evalutate correlation between parameters
#install.packages("corrplot)")
#require(corrplot)
#cor1 <- cor(fre.dat.final[,3:ncol(fre.dat.final)])

```

```{r EnvParmTableFRMT}
require(pander)
EnvParmTable <- data.frame(fre.dat$OrganizationIdentifier[fre.dat$CharacteristicName %in% env.parm],
                           fre.dat$CharacteristicName[fre.dat$CharacteristicName %in% env.parm],
                           fre.dat$ResultMeasure.MeasureUnitCode[fre.dat$CharacteristicName %in% env.parm],
                           fre.dat$MethodDescriptionText[fre.dat$CharacteristicName %in% env.parm]
                           ,row.names = NULL)

EnvParmTable <- EnvParmTable[!duplicated(EnvParmTable),]
#EnvParmTable <- EnvParmTable[-c(1,12),]
row.names(EnvParmTable) <- NULL
colnames(EnvParmTable) <- c("Organization","Characteristic","Unit of Measure","Method URL")

set.caption("Selected variables for analysis")
set.alignment("left")
pander(EnvParmTable)
```